
---
title: "Resources"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: yes
    css: assets/styles.css
    includes:
      after_body: footer.html
      in_header: header.html
---

```{r tabfig, echo=FALSE}
source("assets/tabfig.R")
```

## Books {#books}

### Main texts {#texts}

```{r do-books, echo=FALSE}
width <- "180px"
tab(class="cellpadding", width="800px",
  tr(
  	tabfig("FriendlyMeyer", "icons/ddar-cover.png", 
  	       "https://www.taylorfrancis.com/books/mono/10.1201/b19022/discrete-data-analysis-michael-friendly-david-meyer", width=width),
  	tabtxt("Friendly & Meyer (2016),",  a("Discrete Data Analysis with R: Visualization and Modeling Techniques for  
  	Categorical and Count Data",
  	                         href="https://www.taylorfrancis.com/books/mono/10.1201/b19022/discrete-data-analysis-michael-friendly-david-meyer"),
  	       ". This book provides the syllabus and main content for the course.", 
  	       "Use the code ADC22 for a 30% discount from the publisher's web site.", br(),
           "Web site for the book:", aself("http://ddar.datavis.ca/"))
  	),
  tr(
  	tabfig("Agresti", "icons/agresti-intro-3rd.jpg", 
  	       "https://www.wiley.com/en-us/An+Introduction+to+Categorical+Data+Analysis%2C+3rd+Edition-p-9781119405283", width=width),
  	tabtxt("Alan Agresti (2019),",  a("Introduction to Categorical Data Analysis, 3rd Ed.",
  	                         href="https://www.wiley.com/en-us/An+Introduction+to+Categorical+Data+Analysis%2C+3rd+Edition-p-9781119405283"),
  	       ". A somewhat parallel book, offering a different perspective on categorical data analysis.", br(),
           "An ebook version can be purchased from the York Bookstore. See ",
  	       a("How to purchase your course materials for PSYC 6136", href="papers/YorkBookstore-PSYC6136.docx"))
  	)
)
```

### Supplementary readings {#supp}

```{r do-books2, echo=FALSE}
width <- "180px"
tab(class="cellpadding", width="800px",
  tr(
  	tabfig("Agresti", "icons/agresti-cover-3rd.jpg", 
  	       "https://www.wiley.com/en-us/Categorical+Data+Analysis,+3rd+Edition-p-9780470463635", width=width),
  	tabtxt("Agresti (2013)",  a("Categorical Data Analysis",
  	       href="https://www.wiley.com/en-us/Categorical+Data+Analysis,+3rd+Edition-p-9780470463635"),
  	       ". A much more technical book, that many consider the 'bible' for categorical data analysis methods.", br(),
  	       a("Web site for the book", "http://www.stat.ufl.edu/~aa/cda/cda.html"), br(),
  	       a("Solutions manual for R", "https://home.comcast.net/~lthompson221/Splusdiscrete2.pdf"), br(),
  	       "A PDF copy of this book is available to students in this course."
           )
  	),
  tr(
  	tabfig("Fox", "icons/Fox-APR3rd.jpg", 
  	       "https://us.sagepub.com/en-us/nam/applied-regression-analysis-and-generalized-linear-models/book237254", width=width),
  	tabtxt("Fox (2015)",  a("Applied Regression Analysis and Generalized Linear Models",
  	       href="https://us.sagepub.com/en-us/nam/applied-regression-analysis-and-generalized-linear-models/book237254"),
  	       ". An excellent text on linear models; Part IV on Generalized Linear Models provides a clear and comprehensive discussion.")
  	)
  )
```

## Software {#software}

In lectures and lab sessions I will be using [R](http://www.r-project.org)
software exclusively, together with the [R Studio](http://www.rstudio.com/)
user interface for R.

You are well-advised to download and install these to your computer so you can follow along.

* Instructions to install R and R Studio for [Windows](tutorials/installing-r-rstudio-windows.pdf) and [Mac](tutorials/installing-r-rstudio-mac.pdf). 
There is also a [learnr tutorial](https://learnr-examples.shinyapps.io/ex-setup-r/)
that guides you through the steps.

* My [install-vcd-pkgs.R](R/install-vcd-pkgs.R) R script, to install the most useful packages for this course. Download and run this in your R or RStudio console.

* [R Studio cheatsheets](tutorials/rstudio-cheat-sheets.pdf) A handy collection of cheat sheets for R, R Studio and a number of the most useful R packages. You can also get some of these in R Studio from the menu `Help -> Cheat sheets`.

* [An Introduction to R Graphics](http://www.datavis.ca/courses/RGraphics/) Notes from my SCS short course on R Graphics. 

* The [`vcdExtra` package](https://friendly.github.io/) contains a number of vignettes, each giving practical methods for working with categorical data and models.

  + [Creating and manipulating frequency tables](https://friendly.github.io/vcdExtra/articles/creating.html)
  + [Tests of Independence](https://friendly.github.io/vcdExtra/articles/tests.html)
  + [Mosaic plots](https://friendly.github.io/vcdExtra/articles/mosaics.html)
  + [Loglinear models](https://friendly.github.io/vcdExtra/articles/loglinear.html)

* Short [Making tables in R](tutorials/tables.html) tutorial. Also: Udi's [Psy3136 Tables](tutorials/Tables-brief.html) tutorial on making tables with the `rempsyc` and `apaTables` packages.
  
### Making plots with ggplot2 {#ggplot}

The majority of the graphs in [DDAR](https://www.taylorfrancis.com/books/mono/10.1201/b19022/discrete-data-analysis-michael-friendly-david-meyer) and in my lectures
use custom graphic methods implemented in the `vcd`, `vcdExtra`, `ca` and other packages
specific to categorical data analysis.
Yet it is also helpful to learn
how to make and customize graphs using `ggplot2`, the modern _lingua franca_ 
for specifying and producing graphs.

Here are a few links that I find useful:

* The  [ggplot2 Book](https://ggplot2-book.org/) is the bible of graphing
with the _Grammar of Graphics_ framework. It explains the logic of layers,
geoms, scales, themes, etc.  It has many, many small, easily understood examples.

* The [R Graphics Cookbook](https://r-graphics.org/) is a great resource of recipes for ggplot2, organized by type of graph (bar charts, line graphs, scatterplots, ...) and things you might want to do with them.

* My favorite book on general ideas about graphs:
Claus Wilke, [Fundamentals of Data Visualization](https://clauswilke.com/dataviz/). 
Well thought out, a wide range of topics, good practical advice, lots of examples, but no R code in the book.

* My lecture slides from [Psy 6135: Psychology of Data Visualization](https://friendly.github.io/6135/) give a reasonable overview:
	+ [Introduction to ggplot2](https://friendly.github.io/6135/lectures/ggplot-intro.pdf)
	+ [ggplot2: Going further in the tidyverse](https://friendly.github.io/6135/lectures/ggplot-further.pdf)


### Papers, talks, blogs and others {#papers}

* Friendly (2002). [A brief history of mosaic displays](https://www.datavis.ca/papers/mosaic-history.pdf), _JCGS_ **11**(1), 89-107.
Traces the origin of visual and conceptual ideas leading to modern
mosaic displays.

* Slides from a talk at CARME 2011, [Advances in Visualizing Categorical Data Using the vcd, gnm and vcdExtra Packages in R](https://www.datavis.ca/papers/adv-vcd-4up.pdf)


* Slides from a talk at CARME 2015, [General Models and Graphs for Log Odds and Log Odds Ratios](https://www.datavis.ca/papers/CARME2015-2x2.pdf). This re-considers some standard loglinear models and graphical methods (correspondence analysis, mosaic plots) from the
perspective of models and visualization for log odds and log odds ratios.



